In-Band Synchronization for\\Distributed SDN Control Planes\thanks{Supported by the European Institute of Innovation \& Technology (EIT) project 13153 (Software Defined Networking) and the FP7 EU project UNIFY.}

### 2015-11-18 ccr-camera/inband-sync.tex
cr format
done

%\documentclass[11pt,pdftex,letter]{article}
\documentclass[conference]{sigcomm-alternate}

@@

\begin{figure}[t]
\centering
\includegraphics[width=.56\columnwidth]{loadbal-1.pdf}~\hspace{-.7cm}~\includegraphics[width=.56\columnwidth]{loadbal-2.pdf}\\
\caption{\small \emph{Left:} Without synchronization, the two controllers naturally choose to
install their flows on the left link, which results in an undesirable
unbalanced state.
 \emph{Right:}
 With synchronization, i.e., by bundling the flow installation with a Compare-and-Set (CAS) primitive
 depending on the policy id (\emph{\pid}), this problem is avoided.}\label{fig:CAS-example}
\end{figure}

@@

The meta-configuration is implemented as a set of flow entries
designed in a way that it does not affect the processing of data plane
traffic. 
We will later discuss later how this can be achieved, for example in 
today's OpenFlow.

%One way to achieve this, for example, is to set a low priority to
%the meta flow entries and ensure that the regular flow entries
%maintain higher priority levels. We will discuss other options later
%in the context of OpenFlow.

%\petr{How to make sure that traffic is not affected by the ``meta'' part?}
\vspace{1mm}
\noindent\textbf{Synchronization.}
 %
We provide the controllers with the ability to execute
sequences of \emph{operations} on the
switch configuration in an \emph{atomic} (transactional) or \emph{non-atomic} mode.
In the atomic mode, either \emph{all} operations in the sequence, henceforth simply
called \emph{transaction},
should be executed, or none (the transaction is rejected).
In the non-atomic mode, no atomicity guarantees are provided.

We introduce two constructs
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\exec\{op_1,\;\ldots\;op_k\}$ and $\execatomic\{op_1,\;\ldots\;op_k\}$
which perform a sequence of commands $op_1,\ldots,op_k$ in
a regular and atomic way, respectively.
In particular, the construct $\execatomic\{op_1,\;\ldots\;op_k\}$  
returns $\ack$ if and only if all commands in the sequence are performed successfully; otherwise
$\error(\xid,\ecode)$ is returned, providing
the identifier of the first command in the sequence that failed
together with the corresponding error code.

@@

Therefore, we provide the following (transactional) operations.
A $\textit{write}(\addr,k)$ operation sets the content at memory location
$\addr$ to $k$.
A $\textit{\compare}(\addr,k)$ operation \emph{aborts} the transaction
if and only if the content at
$\addr$ is \emph{not} $k$.
Intuitively, by combining compare and write with a set of regular
switch update commands (e.g., FlowMod in OpenFlow) 
in a transaction  we can make sure that the
configuration is changed and the current policy identifier is updated
only if the check operation \emph{succeeds} (i.e., does not abort).

@@

To read the current configuration of a switch (the existing flow entries), the controller should send
the OpenFlow \emph{add flow monitor} command with the \textsf{ofpt\_initial} optional flag.
In this paper, we write \textbf{read-state} for sending
this message, receiving a response, and returning the received configuration.
\petr{Is it atomic? Should we say that the command returns a ``snapshot'' of the
  configuration?}

@@

We now describe Algorithms~\ref{alg:claim}-\ref{alg:compare} in more
detail.
Each of the operations generates one or two FlowMod commands.
The \textit{match} part in these commands specifies only the $64$bit meta-data
field in the header space, leaving the remaining fields at default
values.
In defining the value and the mask,
% of the math part,
we
use the notation $a\concat b$ to represent a concatenation of two
 integers $a$ and $b$: $a\concat b := (a<<s)+b$, where $s$ is the size of $b$ in bits.
We also write \texttt{$\ones{s}$} (\texttt{$\zeros{s}$}) for  a $s$ bit 
string of $1$'s ($0$'s).
%
The action part in these commands is defined as an integer, which in
practice can be implemented by a \emph{set meta-data} instruction where the written value is that integer.

@@

\vspace{1mm}
\noindent\textbf{Write and compare.}
The memory \memwrite and \compare
 operations are implemented using similar techniques.
To write a value $k$ in a memory address $\addr$ (Algorithm~\ref{alg:write}), a new flow entry is
added, in which the value part of the match field contains $\addr$ and
the mask part contains $k$. 
Therefore, the effective match string od the entry 
 is $k'\concat \addr$, where $k'$ is result os replacing every zero bit in $k$ with a wildcard ($*$).
  In order to rewrite a memory location, the old entry needs to be deleted and a new one need to be added atomically.
  %Values written to the same address will replace one another,
%following the command definition to replace the old flow entry with a new one in case they
%share the same match parts.
Similarly to claim entries, we set the action to a magic value,
$\memmagic$, in order to distinguish them from policy entries.
%[[PK not needed?
%returned by \textbf{read-state}.
%]]

To check whether a value $k$ is written at a given
address $\addr$ (Algorithm ~\ref{alg:compare}),
a controller tries to add a new entry the same way as in the \memwrite operation,
but additionally setting the \texttt{\checko} flag. 
If a different value $k'$ is currently written, 
%our action differs from the existing entry, but 
the entry inflicts a failure, as it overlaps but does not coincide
with the existing entry.
Indeed, the value parts of the two entries are identical, but the
masks differ.
% \liron{(different mixtures of ones and wildcards in the $k'$ part)},
%thereby inflicting a failure. 
In case the currently written value is $k$, the existing flow entry is replaced
with an identical one.

\liron{Note that comparing an uninitialized address always succeeds and also writes to it the compared value. Initializing required addresses can be safely performed by an a-priori comparison with zero.}
\petr{A bit unclear}

@@

Let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (\pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The latter ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
Algorithm~\ref{alg:simple-update} can therefore also be used
to render a na\"ive, single-controller load-balancing algorithm
consistent, without any modifications in the update function.

@@

{\scriptsize
\begin{algorithm}[t]
    \caption{Policy update with only CAS}
    \label{alg:simple-update}
    \begin{algorithmic}[1]
    \Require policy update function \textsc{\ufunc}, policy id address $\paddr$
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
 			\State $\pid,\textit{policy}\gets$ \textbf{read-state} %switch configuration}
 			\State $\textit{update\_cmds}\gets \textsc{\ufunc}(policy)$
 			\startTxn
	 			\State $CAS(\paddr,\pid,\pid+1)$,
	 			\State $\textit{update\_cmds}$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
     		%\State $res \gets $ commit transaction
     	\Until{$\textit{res}=\ack$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return $\textit{res}$

@@

A less trivial Algorithm~\ref{alg:update} adresses the problem of
policy identifiers growing without bound in  
Algorithm~\ref{alg:simple-update} by, additionally, using our id-claimer abstraction.
%to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and 
claims it, which prevents another controller from using the id 
for a different policy (Lines~\ref{update:read-state1}-\ref{update:read-state2}).Then the controller claims the first \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}).
Finally, within an atomic transaction (\textbf{\execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then updates the
policy.
%[[PK 
%We observe that our  algorithm is only
%\emph{lock-free}, as it allows scenarios in which only one live controller
%succeeds in installing its updates. We can easily make it
%wait-free by reading the state of the switch and claiming the current
%policy id within the same {\execatomic} construct.
%]]

\vspace{1mm}
\noindent\textbf{A note on efficiency.}

We observe that our implementations of synchronization primitives
(Algorithms~\ref{alg:claim}-\ref{alg:compare}) 
%is \emph{switch-space efficient}
%in the sense that 
require only a constant number of flow entries per memory address.
The number of used addresses is constant for  
Algorithm~\ref{alg:simple-update} and linear in the number of
controllers for Algorithm~\ref{alg:update}. 
%[[PK not constant: proportional to the number of failures. But this
%is kind of obvious.
%Also, in case of controller failures, at most a constant amount of space is lost.
%]]

\stefan{We also note that while we described our scheme in a simplified
way where numbers continuously increase, it is straight-forward to use a modulo
scheme. While in our case 32-bit numbers are used, 
the flow table width (number of bits which can be matched in parallel)
of modern switches is 40-bits and more.}

 \liron{I think we should say something like: "... of modern switches can get as high as a few hundreds'. }

\petr{The discussion is very confusing. Do we need it?}

@@

To investigate the feasibility of our approach,
we implemented a proof-of-concept of our algorithms
as a python library which wraps ovs-ofctl, 
a well-known command line tool for monitoring and administering OpenFlow switches. 
The code is publicly available at \url{https://github.com/lironsc/of-sync-lib}.

@@

For instance, we experimented with a setting where three processes
(controllers) issue thousands of concurrent policy updates
to the switch, 
each update consisting in replacing a single rule with another one.
%[[PK unclear
%When verifying the transactions, 
%]]
We could confirm that 
despite the concurrency, 100 \% of the transactions have been
eventually completed.
Of course, due to the concurrency, some transactions also needed to abort
because of conflicts. 
\petr{Reads strange: 100\% successfullt complete and some transactions
  abort...}

While our framework can be used as a synchronization tool itself, we 
currently also work on implementing it as an integral part of existing controllers 
such as Ryu. 
\petr{Did not get the difference between ``itself'' and ``part of
  existing controllers''.}

@@

We believe that our centralized in-band
synchronization approach is natural and attractive, as it avoids the
complexities and limitations of distributed out-of-band agreement protocols, which may 
introduce additional points of potential failures.
Moreover, the switch is the element implementing the controller policies anyway.
The mechanisms presented in this paper are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{stn,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions~\cite{pyretic}.
In general, we see our paper as a first step, and more
work is required to investigate and compare useful
in-band synchronization abstractions, also in terms of performance.

>>>>

%\documentclass[11pt,pdftex,letter]{article}

%\documentclass[conference]{sigcomm-alternate}

%\documentclass[dvipdfm]{sig-alternate-2006}

\documentclass{sig-alternate-2006}

@@

\begin{figure}[t]
\centering
\includegraphics[width=.56\columnwidth]{loadbal-1.pdf}~\hspace{-.7cm}
~\includegraphics[width=.56\columnwidth]{loadbal-2.pdf}\\
\caption{\small \emph{Left:} Without synchronization, the two controllers naturally choose to
install their flows on the left link, which results in an undesirable
unbalanced state.
 \emph{Right:}
 With synchronization, i.e., by bundling the flow installation with a Compare-and-Set (CAS) primitive
 depending on the policy id (\emph{\pid}), this problem is avoided.}\label{fig:CAS-example}
\end{figure}

@@

The meta-configuration is implemented as a set of flow entries
designed in a way that it does not affect the processing of data plane
traffic. 
We will discuss later how this can be achieved, for example, in 
today's OpenFlow. 
%One way to achieve this, for example, is to set a low priority to
%the meta flow entries and ensure that the regular flow entries
%maintain higher priority levels. We will discuss other options later
%in the context of OpenFlow.

%\petr{How to make sure that traffic is not affected by the ``meta'' part?}
\vspace{1mm}
\noindent\textbf{Synchronization.}
 %
We provide the controllers with the ability to execute
sequences of \emph{operations} on the
switch configuration in an \emph{atomic} or \emph{non-atomic} mode.
In the atomic mode, either \emph{all} operations in the sequence, henceforth simply
called \emph{transaction},
should be executed, or none (the transaction is rejected).
In the non-atomic mode, no atomicity guarantees are provided.
%\liron{all these jumps from atomic to transactional and back is confusing...}
%

%For this purpose, we introduce two constructs:
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
%$\exec\{\ldots\}$ and
%$\execatomic\{\ldots\}$.
In particular, construct $\exec\{op_1,\;\ldots\;op_k\}$ is used to perform a sequence of commands $op_1,\ldots,op_k$ in
a conventional way.
Construct $\execatomic\{op_1,\;\ldots\;op_k\}$ ensures that either all
commands are successfully performed, in which case it returns
$\ack$, or, if some of the commands cannot be performed, aborts the
whole sequence and returns
$\error(\xid,\ecode)$, providing
the identifier of the first command in the sequence that failed
together with the corresponding error code.

@@

Therefore, we provide the following transactional operations.
A $\textit{write}(\addr,k)$ operation sets the content at memory location
$\addr$ to $k$.
A $\textit{\compare}(\addr,k)$ operation \emph{aborts} the transaction
if and only if the content at
$\addr$ is \emph{not} $k$.
Intuitively, by combining compare and write with a set of regular
switch update commands (e.g., FlowMod in OpenFlow) 
in a transaction  we can make sure that the
configuration is changed and the current policy identifier is updated
only if the check operation \emph{succeeds} (i.e., does not abort).

@@

To read the current configuration of a switch (the existing flow entries), the controller should send
the OpenFlow \emph{add flow monitor} command with the \textsf{ofpt\_initial} optional flag.
In this paper, we write \textbf{read-state} for sending
this message, receiving a response, and returning the received configuration.
%\petr{Is it atomic? Should we say that the command returns a ``snapshot'' of the
%  configuration?}
%\liron{I am not sure if this is atomic, considering that we execute other things in a bundle then this is practically atomic for us. Perhaps we can leave it unspecified as it is now?}
%[[PK moved this discussion to the algorithms 6 and 7]]

@@

We now describe Algorithms~\ref{alg:claim}-\ref{alg:compare} in more
detail.
Each of the operations generates one or two FlowMod commands.
The \textit{match} part in these commands specifies only the $64$bit meta-data
field in the header space.
%, leaving the remaining fields at default
%values.
%\liron{this is not accurate, the match part simply considers only the metadata field. no need to discuss other fields, in any case matching doesn't change fields and doesn't assume any values to other fields.}
In defining the value and the mask,
% of the math part,
we
use the notation $a\concat b$ to represent a concatenation of two
 integers $a$ and $b$: $a\concat b := (a<<s)+b$, where $s$ is the size of $b$ in bits.
We also write \texttt{$\ones{s}$} (\texttt{$\zeros{s}$}) for  a $s$ bit 
string of $1$'s ($0$'s).
%
The action part in these commands is defined as an integer, which in
practice can be implemented by a \emph{set meta-data} instruction where the written value is that integer.

@@

\vspace{1mm}
\noindent\textbf{Write and compare.}
The memory \memwrite and \compare
 operations are implemented using similar techniques.
To write a value $k$ in a memory address $\addr$ (Algorithm~\ref{alg:write}), a new flow entry is
added, in which the value part of the match field contains $\addr$ and
the mask part contains $k$. 
Therefore, the effective match string of the entry 
 is $k'\concat \addr$, where $k'$ is result of replacing every zero bit in $k$ with a wildcard ($*$).
  In order to rewrite a memory location, the old entry needs to be deleted and a new one need to be added atomically.
  %Values written to the same address will replace one another,
%following the command definition to replace the old flow entry with a new one in case they
%share the same match parts.
Similarly to claim entries, we set the action to a magic value,
$\memmagic$, in order to distinguish them from policy entries.
%[[PK not needed?
%returned by \textbf{read-state}.
%]]

To check whether a value $k$ is written at a given
address $\addr$ (Algorithm ~\ref{alg:compare}),
a controller tries to add a new entry the same way as in the \memwrite operation,
but additionally setting the \texttt{\checko} flag. 
If a different value $k'$ is currently written, 
%our action differs from the existing entry, but 
the entry inflicts a failure, as it overlaps but does not coincide
with the existing entry.
Indeed, the value parts of the two entries are identical, but the
masks differ.
% \liron{(different mixtures of ones and wildcards in the $k'$ part)},
%thereby inflicting a failure. 
In case the currently written value is $k$, the existing flow entry is replaced
with an identical one.
%[[PK on initialization
Before using an address $\addr$, it is advised that a controller safely
initializes it with a special initial value $0$ by calling
$\textit{\compare}(\addr,0)$: 
the first such command succeeds, and any 
subsequent one will fail. 
%]]

%\liron{Note that comparing an uninitialized address always succeeds and also writes to it the compared value. Initializing required addresses can be safely performed by an a-priori comparison with zero.}
%\petr{A bit unclear}

%\liron{SECOND TRY: Note that comparing an address for the first time, without writing it first, always succeeds and also writes to it the compared value. In case where such comparison may occur to some address $addr$, it is advised to first perform a safe initialization by calling $\textit{\compare}(\addr,0)$ which fails in case that $addr$ has already been used (and the initialization is not desired).}

@@

Let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (\pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The CAS operation ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
Note that, since all policy modifications are executed within a
bundle providing controller-atomicity, the \textbf{read-state} command
in Line~\ref{line:readstate} appears atomic.  
Algorithm~\ref{alg:simple-update} can therefore also be used
to render a na\"ive, single-controller load-balancing algorithm
consistent, without any modifications in the update function.

@@

{\scriptsize
\begin{algorithm}[t]
    \caption{Policy update with only CAS}
    \label{alg:simple-update}
    \begin{algorithmic}[1]
    \Require policy update function \textsc{\ufunc}, policy id address $\paddr$
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
 			\State $\pid,\textit{policy}\gets$ \textbf{read-state} \label{line:readstate}%switch configuration}
 			\State $\textit{update\_cmds}\gets \textsc{\ufunc}(policy)$
 			\startTxn
	 			\State $CAS(\paddr,\pid,\pid+1)$,
	 			\State $\textit{update\_cmds}$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
     		%\State $res \gets $ commit transaction
     	\Until{$\textit{res}=\ack$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return $\textit{res}$

@@

Algorithm~\ref{alg:simple-update} is correct assuming that policy
identifiers may grow without bound. 
In practice, we can use a module scheme on the 32-bit field of a flow entry
used for storing the policy identifier, which, in principle,
may lead to inconsistencies when the relative speeds of controllers is
very high.   
Algorithm~\ref{alg:update} adresses this issue 
by using our id-claimer abstraction.
%to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and 
claims it, which prevents another controller from using the id 
for a different policy
(Lines~\ref{update:read-state1}-\ref{update:read-state2}). 
Then the controller selects any \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}). %\liron{I have replaced "claims the first" with "chooses an" following Petr email}
Finally, within an atomic transaction (\textbf{\execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then tries to update the
policy.
%

%[[PK 
%We observe that our  algorithm is only
%\emph{lock-free}, as it allows scenarios in which only one live controller
%succeeds in installing its updates. We can easily make it
%wait-free by reading the state of the switch and claiming the current
%policy id within the same {\execatomic} construct.
%]]

\vspace{1mm}
\noindent\textbf{A note on efficiency.}
%
We observe that our implementations of synchronization primitives
(Algorithms~\ref{alg:claim}-\ref{alg:compare}) 
%is \emph{switch-space efficient}
%in the sense that 
require only a constant number of flow entries per memory address or claim.
The number of used addresses is constant for  
Algorithm~\ref{alg:simple-update} and linear in the number of
controllers for Algorithm~\ref{alg:update}. 
%[[PK not constant: proportional to the number of failures. But this
%is kind of obvious.
%Also, in case of controller failures, at most a constant amount of space is lost.
%]]

%\stefan{We also note that while we described our scheme in a simplified
%way where numbers continuously increase, it is straight-forward to use a modulo
%scheme. While in our case 32-bit numbers are used, 
%the flow table width (number of bits which can be matched in parallel)
%of modern switches is 40-bits and more.}

 %\liron{I think we should say something like: "... of modern switches can get as high as a few hundreds'. }

%\petr{The discussion is very confusing. Do we need it?}

@@

To investigate the feasibility of our approach,
we implemented a proof-of-concept of our algorithms
as a python library which wraps ovs-ofctl, 
a well-known command line tool for monitoring and administering OpenFlow switches. 
The code is available online at
\url{https://github.com/lironsc/of-sync-lib}. 
%\liron{perhaps we can put it as reference? (putting the link in bib)}

@@

For instance, we experimented with a setting where three processes
(controllers), concurrently issuing $1000$ concurrent policy updates
each, where every update replaces a single rule with another one.
%[[PK unclear
%When verifying the transactions, 
%]]
We could confirm that 
despite the concurrency, 100\% of the updates 
eventually completed in committing their transactions, possibly after
several aborts caused by conflicts. 
%Of course, due to the concurrency, some transactions also needed to abort
%because of conflicts. 
%\petr{Reads strange: 100\% successfullt complete and some transactions
 % abort...}

@@

We believe that our centralized in-band
synchronization approach is natural and attractive, as it avoids the
complexities and limitations of distributed out-of-band agreement protocols, which may 
introduce additional points of potential failures.
Moreover, the switch is the element implementing the controller policies anyway.
The mechanisms presented in this paper are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{stn,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions~\cite{pyretic}.
We see our paper as a first step, and more
work is required to investigate and compare useful
in-band synchronization abstractions, also in terms of performance.
While our framework can be used as a synchronization library, we 
also work on integrating it into an existing SDN controller, e.g., Ryu.
